Supermartingales and Termination Analysis of Statistical PCF

### 2020-10-13 probabilisticProgrammingMartingales.tex
Prove Theorem 2, thm:CbvIsTerminatingest.
Because of the restrictions on when sample reductions can occur,
this isn't actually following the same logic as the standard theorem
that lazy evaluation terminates if anything does, but it was pretty
simple anyway.

\paragraph{}
\begin{lemma}
The relation $\Rightarrow$ is Church-Rosser.
\end{lemma}
\begin{proof}
Suppose that $(M,s) \Rightarrow (M_1,s_1)$ and also $(M,s) \Rightarrow (M_2,s_2)$, then it is required to prove that there is some $(M',s')$ such that both $(M_1,s_1) \Rightarrow^* (M',s')$ and $(M_2,s_2) \Rightarrow^* (M',s')$. Let the positions of the redexes in $M \to M_1$ and $M \to M_2$ be $\alpha_1$ and $\alpha_2$ respectively.

@@

\begin{theorem} \label{thm:CbvIsTerminatingest}
If $M$ terminates with some reduction strategy $f$ and samples $s$, it terminates with $\cbv$ and $s$.
\end{theorem}
\begin{proof}
\end{proof}

>>>>

\paragraph{}
\begin{lemma} \label{churchRosser}
The relation $\Rightarrow$ is Church-Rosser.
\end{lemma}
\begin{proof}
Suppose that $(M,s) \Rightarrow (M_1,s_1)$ and also $(M,s) \Rightarrow (M_2,s_2)$, then it is required to prove that there is some $(M',s')$ such that both $(M_1,s_1) \Rightarrow^* (M',s')$ and $(M_2,s_2) \Rightarrow^* (M',s')$. Let the positions of the redexes in $M \to M_1$ and $M \to M_2$ be $\alpha_1$ and $\alpha_2$ respectively.

@@

\begin{theorem} \label{thm:CbvIsTerminatingest}
If $M$ terminates with some reduction strategy $f$ and samples $s$, it terminates with $\cbv$ and $s$.
\end{theorem}
\begin{proof}
  Suppose $M$ terminates with the reduction strategy $f$ and samples $s$. Let $(M,s) = (M_0,s_0) \Rightarrow_f \dots \Rightarrow_f (M_n,s_n)$, where $M_n$ is a value. By Lemma \ref{canonicalCousins}, the reduction sequence $(Sk(M_i))_i$ is related by $\sim_c^*$ to some skeletal reduction sequence $X$ which is in call-by-value order. For each skeletal reduction sequence $X_j$ in the sequence $(Sk(M_i))_i \sim_c \dots \sim_c X$, it is possible to define a reduction sequence of terms such that $(M,s) = (N_{j,0},s_{j,0}) \Rightarrow \dots \Rightarrow (N_{j,n_j},s_{j,n_j})$ and $Sk(N_{j,k}) = X_{j,k}$, by at each step applying the reduction at the same place as in $X_j$, essentially just filling in the reals in the skeletal reduction sequence to make it a reduction sequence of terms (it will be shown shortly that the correct branch is taken in each $\textsf{if}$-reduction). Assume for induction on $j$ that $(N_{j,n_j},s_{j,n_j}) = (M_n,s_n)$. To prove that this holds true for $j+1$ as well, it suffices to show that the corresponding term and samples immediately after the reductions involved in this $\sim_c$ step match, as from that point on, the $\Rightarrow$-sequences will match, as they start from the same point and have all the same reduction positions. Letting the term and remaining samples immediately before the reductions involved in this $\sim_c$ step be $(N,t)$ (which is the same in both $N_j$ and $N_{j+1}$ for the same reason again, they start at the same point and the sequences of reduction positions up to this point are the same), and the terms and samples after these reductions be $(O_1,t_1)$ and $(O_2,t_2)$, so that these match the $N$, $O_1$ and $O_2$ in the definition of $\sim_c$, $t_1 = t_2$ because both of them are, by Lemma \ref{lem:sim-M-N}, equal to the restriction of $t$ to the subset of $Rch(N)$ that starts with the reduction sequences to $O_1$ and $O_2$ respectively, and the potential positions in those subsets of $Rch(N)$ are identified with each other by $\sim$ because of the same case of $\sim_c$ that means $N_j$ and $N_{j+1}$ are related. If none of the relevant reductions are $\tsample$-reductions, $O_1 = O_2$ trivially. If there are samples taken (which must be the arbitrary reduction at $\beta$), it is also required that the same samples are taken in both branches. Similarly to Lemma \ref{churchRosser}, the positions of these $\tsample$-reductions in each branch are related by $\sim_p^*$, therefore they correspond to the same sample in $t$, therefore $O_1 = O_2$. The fact that the terms are equal also implies that the same branch of an $\textsf{if}$ statements will be taken.

  \paragraph{}
  There is therefore a reduction sequence $(M,s) = (N_0,s_0') \Rightarrow \dots \Rightarrow (N_m, s_m')$ which ends in a value ($N_m = M_n$) and is in CBV order. It doesn't immediately follow that this is the $\Rightarrow_{\cbv}$ reduction sequence starting from $(M,s)$ because the fact that $(N_i)_i$ is in CBV order only means that the reductions that do take place in this reduction sequence are in the correct order, not that the reduction sequence finishes once it reaches a value, or that there are no other reductions that would come earlier in CBV order that don't occur at all in $(N_i)_i$. Suppose that at some point $j$ in the reduction sequence, the reduction that occurs is not the next reduction in CBV order, but something else, and that the reduction that should be next is at position $\alpha = \cbv(N_i)$, and the position of the actual reduction $N_j \to N_{j+1}$ is $\beta$. Suppose further that $N_j$ is not a value. Either $\alpha = \cdot$, the root position, or the reduction at the root position is dependent in some way on the reduction at $\alpha$ (e.g. $\alpha = \textsf{if}_1$, or $\alpha = \underline{+};@_1$). In any case, all the reductions in the rest of the reduction sequence are after $\alpha$ in CBV order, therefore none of them are $\leq \alpha$, and there continues to be an unreduced redex at $\alpha$, therefore the reduction at the root position never occurs, therefore $N_m$ is not a value, which is a contradiction, therefore at the point (if any) where $(N_i)_i$ departs from the $\Rightarrow_{\cbv}$ reduction sequence starting at $(M,s)$, it has already reached a value, therefore $(M,s) \Rightarrow_{\cbv} \dots$ does reach a value eventually, i.e.~$M$ terminates with the reduction strategy $\cbv$ and samples $s$.
\end{proof}
